---
title: "Data Analysis #2 Version 2 (75 points total)"
author: "Siddikov, Alisher"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(flux)
library(ggplot2)
library(gridExtra)
library(moments)
# library(rockchalk) # base R code replaces requirement for this package
library(car)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("C:\\Users\\asidd\\Desktop\\MSDS\\401 Statistics with R\\9\\Assignment\\mydata.csv", sep = ",")
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

##### Section 1: (5 points)

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a}

RATIO <- mydata$RATIO
SHUCK <- mydata$SHUCK
VOLUME <- mydata$VOLUME

L_RATIO <- log10(mydata$RATIO)
L_SHUCK <- log10(mydata$SHUCK)
L_VOLUME <- log10(mydata$VOLUME)

par(mfrow = c(1, 2))
hist(RATIO)
qqnorm(RATIO)
qqline(RATIO)

library(rockchalk)
rockchalk::skewness(RATIO)
rockchalk::kurtosis(RATIO)
```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b}
par(mfrow = c(1, 2))
hist(L_RATIO)
qqnorm(L_RATIO)
qqline(L_RATIO)

par(mfrow = c(1, 1))
boxplot(L_RATIO ~ mydata$CLASS, xlab = "CLASS", ylab = "L_Ratio")


rockchalk::skewness(L_RATIO)
rockchalk::kurtosis(L_RATIO)
```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
bartlett.test(RATIO ~ CLASS, data = mydata)
bartlett.test(L_RATIO ~ CLASS, data = mydata)
```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: We can see that the log ratio has a better normal distribution with homogeneous variances across age classes. Log ratio (compare to the non-log ratio) skewness and kurtoses are closer to zero, the histogram is somewhat normally distributed. Outliers are somewhat evenly distributed in the boxplots, and Q-Q plot tails/outliers are closer to the Q-Q straight line which means less skewness in QQ plot. We can also see that the Bartlett's test p-value of 0.526 of log ratio is not less than the significance level of 0.05. Bartlett’s test indicates that the variances in the five age classes don’t differ significantly. This means we cannot reject the null hypothesis that the variance is the same across age classes with log ratio and has homogeneity of variances, and there is no evidence to suggest that the variance in the ratio is different across five different age classes. So, log ratio exhibits better conformance to a normal distribution with homogeneous variances across age classes. ***

***Using Bartlett's test, we can reassure ourselves regarding variance homogeneity by comparing the variance of across age classes to determine whether they are drawn from populations with equal variance. Equal variances across populations are called homogeneity of variances. The test has the null hypothesis that the variances are equal and the alternative hypothesis that they are not equal. Please note that the ratio is derived from shuck weight divided by abalone volume.***


##### Section 2 (10 points) ###############################

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}
summary(aov(L_RATIO ~ CLASS + SEX + CLASS:SEX, data = mydata))
summary(aov(L_RATIO ~ CLASS + SEX, data = mydata))
```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: We would like to identify whether either of the two independent variables (class and sex) or their interaction (class:sex) are statistically significant. A two-way ANOVA was conducted that examined the effect of gender and age class level on interest in ratio. There was no statistically significant interaction (Class:Sex) between the effects of gender and age class level on interest in ratio due to higher p-value. However, two independent variables without interaction are statistically significant.***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(aov(L_RATIO ~ CLASS + SEX, data = mydata))
```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: The ANOVA test for treatment tells us that the five age classes and three genders (male, female, and infant) groups aren’t equally effective, but it doesn’t tell us which treatments differ from one another. We can use a multiple comparison procedure TukeyHSD() to determine if the mean difference between specific pairs of the group is statistically significant. The mean difference between A1 and A2 pairs of age class group as well as Female and Male pairs of gender group are not statistically different from each other due to high adjusted p-values. So, these results suggest male and female abalones can be combined into a single category labeled as adults.***


######  Section 3: (10 points) ##################

(3)(a1) We combine "M" and "F" into a new level, "ADULT". (While this could be accomplished using *combineLevels()* from the 'rockchalk' package, we use base R code because many students do not have access to the rockchalk package.) This necessitated defining a new variable, TYPE, in mydata which had two levels:  "I" and "ADULT". 

```{r Part_3a1}
# here we show how to define the new variable TYPE using only base R functions (no need for outside packages)
mydata$TYPE <- character(nrow(mydata))  # initialize the TYPE column as all blanks
for (i in seq(along = mydata$SEX)) {
  mydata$TYPE[i] <- 'I'
  if (mydata$SEX[i] == 'M' || mydata$SEX[i] == 'F') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on definition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
table(mydata$SEX, mydata$TYPE)

```
(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2}
par(mfrow = c(1, 2))
hist(VOLUME[mydata$TYPE == 'I'], xlab = "Volume", ylab = "Frequency", main = "Infant Volume")
hist(VOLUME[mydata$TYPE == 'ADULT'], xlab = "Volume", ylab = "Frequency", main = "Adult Volume")
```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: According to the histogram distributions, infant volume distributions are skewed right and most of the infant volumes are below 300 whereas adult volume is somewhat normally distributed and the most of adult volumes are between 200 and 600. According to the histogram, it may be possible to roughly separate infants from older abalones by looking at volumes.  Such a separation will not be error free. So, there are not going to be any difficulties separating infants from adults based on VOLUME: less than 300 volumes are infants and more than 300 volumes are adults. ***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b}
## par(mfrow = c(1, 2))
## plot(x = VOLUME, y = SHUCK)
## plot(x = L_VOLUME, y = L_SHUCK)

grid.arrange(
  ggplot(mydata, aes(x = VOLUME, y = SHUCK, color = CLASS)) + geom_point(),
  ggplot(mydata, aes(x = L_VOLUME, y = L_SHUCK, color = CLASS)) + geom_point(),
  ggplot(mydata, aes(x = VOLUME, y = SHUCK, color = TYPE)) + geom_point(),
  ggplot(mydata, aes(x = L_VOLUME, y = L_SHUCK, color = TYPE)) + geom_point())

```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: Log-scale informs on relative changes or percent changes (multiplicative), while linear-scale informs on absolute changes (additive). Log 10 shows a percentage change of volume and shuck (meat without a shell) of abalones and has a positive linear relationship. We can see that infants with higher log (shuck) and log (volume) overlap with adults. Infants with older age classes might not have been classified right. We can also see that the age class and type cluster/group better with log scale; the younger age class groups or infants has fewer volumes and less shucked weight. As we know that age is determined by drilling the shell and counting the number of shell rings using a microscope. This is a difficult and time-consuming process. Also, ring clarity can be an issue. So, with this analysis, we can predict age class and types by looking at relative changes in the volume and shucked weight. We can stop investing time and money for counting rings manually.***


######   Section 4: (5 points) ###################################

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. Reclassify the infants in classes A4 and A5 as ADULTS. This reclassification could have been achieved using *combineLevels()*, but only on the abalones in classes A4 and A5. We will do this recoding of the TYPE variable using base R functions. We will use this recoded TYPE variable, in which the infants in A4 and A5 are reclassified as ADULTS, for the remainder of this data analysis assignment. 

```{r Part_4a1}
for (i in seq(along = mydata$TYPE)) {
  if (mydata$CLASS[i] == 'A4' || mydata$CLASS[i] == 'A5') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on redefinition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
cat('\nThree-way contingency table for SEX, CLASS, and TYPE:\n')
print(table(mydata$SEX, mydata$CLASS, mydata$TYPE))


```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}
model <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata)
summary(model)

```

**Essay Question:  Interpret the trend in CLASS level coefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: 6.	Age classes have negative coefficients which mean as age class increases from A3 to A5 the shucked weight will decrease (in percentage) if the volume and type variables held constant. Slightly increasing negative coefficients for age class effectively penalizes older individuals because the relationship log (shuck) and age class relationship aren't linear; it plateaus and the increase observed in log (shuck) weight age class to age class gets smaller at each step - before actually falling a bit at A5. R uses the first level of a factor variable for the baseline model fit, then estimates the coefficients to be the change from that baseline. So, for this model, "intercept -0.7964188*L_VOLUME" would give us the predicted log for an A1 and Adult. Please note that all age classes are significant except for age class A2 (p-value = 0.10). Since A2 is adjacent to the baseline level A1 and Adult, it suggests that age class A2 has not significant predictability for the dependent variable when controlling for the other predictor variables. Adjusted R square is 0.95 which means that the 95% data points are closer to the regression line and there is an almost perfect linear relationship between percentage change of shucked weight and three independent variables.  So, it indicates that 95% of the variation in shucked weight (in log-scale) is accounted for by the independent variables volume, age class, and type. This indicates the model has the potential to be useful for making predictions. ***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: Type infant measures the avearge difference shucked weight between adults and infants, given the same levels of volume and age classes. So, if we take an adult and an infant with the same levels of volume and age class, the infant abalone shuck weighs, on average, 2.1% less than the adult, holding other factors fixed. Yes, it is one of the important predictors in this regression. It is comparing well to other independent varaibles in term of contribution of log (shucked weight) harvesting decisions. ***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

###### Section 5: (5 points) #################################

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5a}
par(mfrow = c(1, 2))
hist(model$residuals, xlab = "Residuals", main = "Histogram of Residuals")
qqnorm(model$residuals, main = "QQ Plot of Residuals")
qqline(model$residuals)

rockchalk::skewness(model$residuals)
rockchalk::kurtosis(model$residuals)

```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b}

mydata$RESIDUALS <- model$residuals

## plot(log10(mydata$VOLUME), mydata$RESIDUALS, col = mydata$CLASS)
## plot(log10(mydata$VOLUME), mydata$RESIDUALS, col = mydata$TYPE)

grid.arrange(
  ggplot(mydata, aes(x = L_VOLUME, y = RESIDUALS, color = CLASS)) + theme_bw() + geom_point(),
  ggplot(mydata, aes(x = L_VOLUME, y = RESIDUALS, color = TYPE)) + theme_bw() + geom_point(),
  ggplot(mydata, aes(x = CLASS, y = RESIDUALS, color = CLASS)) + theme_bw() + geom_boxplot(),
  ggplot(mydata, aes(x = TYPE, y = RESIDUALS, color = TYPE)) + theme_bw() + geom_boxplot())

bartlett.test(RESIDUALS ~ CLASS, data = mydata)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: 9.	Let’s first understand the residual meaning. It is an error of the regression model resulted from the difference between the actual value and the predicted value. According to the scatterplot, the model is not perfectly fit or quite evenly distributed but there is a denser cluster appears to the right and there is wider cluster (mainly infants from 0 to 1 volume log scale) on the left when plotted against log (volume). However, boxplot show that the class and type data distributed evenly and centered to zero. According to Bartlett’s test, there is strong evidence that age classes are drawn from populations with equal variance. ***

-----

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

#### Section 6: (5 points) ########################

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

# prop.infants shows the impact of increasing the volume cutoff for
# harvesting. The following code shows how to "split" the population at
# a 50% harvest of infants.

n.infants <- sum(prop.infants <= 0.5)
split.infants <- round(min.v + (n.infants + 0.5)*delta, 2)  # This estimates the desired volume.
split.infants

n.adults <- sum(prop.adults <= 0.5)
split.adults <- round(min.v + (n.adults + 0.5)*delta, 2)
split.adults

head(volume.value)
head(prop.infants)
head(prop.adults)
```

(6)(b) Present a plot showing the infant proportions and the adult proportions versus volume.value. Compute the 50% "split" volume.value for each and show on the plot.   

```{r Part_6b}
### rewrite
## plot(x = volume.value, y = prop.infants)
## plot(x = volume.value, y = prop.adults)

## did not produce legend
## ggplot() + 
##  geom_line(aes(x = volume.value, y = prop.infants), color = "red", show.legend = TRUE) +
##  geom_line(aes(x = volume.value, y = prop.adults),  color = "blue", show.legend = TRUE) +
##  geom_vline(xintercept = split.infants, color = "orange", size = 1) +
##  geom_vline(xintercept = split.adults,  color = "orange", size = 1) +
##  geom_hline(yintercept = 0.5, color = "orange", size = 1) +
##  annotate("text", label = paste(split.infants, "\nInfants"), x = split.infants + 50, y = 0.4) +
##  annotate("text", label = paste(split.adults, "\nAdults"), x = split.adults + 50, y = 0.4) +
##  scale_colour_manual(labels = c("Adult", "Infant"), values = c("blue", "red")) +
##  labs(x = "Volume", y = "Proportion", title = "Adult and Infant Proportions") +
##  theme(legend.title = element_blank()) # suppresses legend title


df <- data.frame(type = rep(c("infant", "adult"), each = 10000),
prop = c(prop.infants, prop.adults),
volume = c(volume.value, volume.value))

ggplot(data = df, mapping = aes(x = volume, y = prop, color = type)) +
  geom_line() +
  geom_vline(xintercept = split.infants, color = "orange", size = 1) +
  geom_vline(xintercept = split.adults, color = "orange", size = 1) +
  geom_hline(yintercept = 0.5, color = "orange", size = 1) +
  annotate("text", label = paste(split.infants, "\nInfants"), x = split.infants + 50, y = 0.4) +
  annotate("text", label = paste(split.adults, "\nAdults"), x = split.adults + 50, y = 0.4) +
  scale_colour_manual(labels = c("Adult", "Infant"), values = c("blue", "red")) +
  labs(x = "Volume", y = "Proportion", title = "Adult and Infant Proportions") +
  theme_bw() +
  theme(legend.title = element_blank()) # suppresses legend title 


```

**Essay Question:  The two 50% "split" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The values suggest a good cutoff (threshold) points for harvesting between the splits. We identified the volume thresholds at which 50% of the infant abalones at or above 133.82 volumes and 50% of the adult abalones at or above 384.51 volumes can be harvested. ***

-----

This part will address the determination of a volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

###### Section 7: (10 points)  #######################

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}
diff <- ((1 - prop.adults) - (1 - prop.infants))
head (diff)

ggplot() + 
  geom_line(aes(x = volume.value, y = diff), color = "red") +
  geom_vline(xintercept = split.infants, color = "orange", size = 1) +
  geom_vline(xintercept = split.adults, color = "orange", size = 1) +
  annotate("text", label = paste(split.infants, "\nInfants"), x = split.infants - 50, y = 0.55) +
  annotate("text", label = paste(split.adults, "\nAdults"), x = split.adults + 50, y = 0.55) +
  labs(x = "Volume", y = "Proportion Difference", title = "Difference in Harvest Proportions") +
  theme_bw()

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

```{r Part_7c}
max <- round(volume.value[which.max(smooth.difference)], 3)

ggplot() + 
  geom_line(aes(x = volume.value, y = diff), color = "red") +
  geom_line(aes(x = volume.value, y = smooth.difference), color = "blue") +
  geom_vline(xintercept = max, color = "orange", size = 1) +
  annotate("text", label = paste("volume =", max), x = max + 20, y = 0.4, angle = 90) +
  labs(x = "Volume", y = "Proportion Difference", title = "Difference in Harvest Proportions") +
  theme_bw()

```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions (NOTE:  the adult harvest proportion is the "true positive rate" and the infant harvest proportion is the "false positive rate").

Code for calculating the adult harvest proportion is provided.

```{r Part_7d}
cutoff0 <- max
cutoff0

cutoff0.a <- (1 - prop.adults)[which.max(smooth.difference)]
cutoff0.i <- (1 - prop.infants)[which.max(smooth.difference)]

paste("True positive rate: ", cutoff0.a) # [1] 0.7416332
paste("False positive rate:", cutoff0.i)


```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

######  Section 8: (10 points)  ###################

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions.

```{r Part_8a}
cutoff1 <- volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786
cutoff1

cutoff1.a <- (1 - prop.adults)[which(volume.value == cutoff1)]
cutoff1.i <- (1 - prop.infants)[which(volume.value == cutoff1)]

paste("True positive rate: ", cutoff1.a)
paste("False positive rate:", cutoff1.i)
```

(8)(b) Another cutoff is one for which the proportion of adults not harvested equals the proportion of infants harvested. This cutoff would equate these rates; effectively, our two errors:  'missed' adults and wrongly-harvested infants. This leaves for discussion which is the greater loss:  a larger proportion of adults not harvested or infants harvested?  This cutoff is 237.7383. Calculate the separate harvest proportions for infants and adults using this cutoff. Show these proportions.  Code for determining this cutoff is provided.  

```{r Part_8b}
cutoff2 <- volume.value[which.min(abs(prop.adults - (1-prop.infants)))] # [1] 237.6391
cutoff2

cutoff2.a <- (1 - prop.adults)[which(volume.value == cutoff2)]
cutoff2.i <- (1 - prop.infants)[which(volume.value == cutoff2)]

paste("True positive rate: ", cutoff2.a)
paste("False positive rate:", cutoff2.i)

```


##### Section 9: (5 points) ###########

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (7) and (8) on this plot and label each. 

```{r Part_9}
ggplot() + 
  geom_line(aes(x = (1 - prop.infants), y = (1 - prop.adults)), color = "blue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = 2) +
  geom_point(aes(cutoff0.i, cutoff0.a), shape = 21, size = 3) +
  geom_point(aes(cutoff1.i, cutoff1.a), shape = 21, size = 3) +
  geom_point(aes(cutoff2.i, cutoff2.a), shape = 21, size = 3) +
  annotate("text", label = paste("max. difference\n", round(cutoff0, 1)), x = cutoff0.i + 0.08, y = cutoff0.a - 0.08) +
  annotate("text", label = paste("equal harvest\n", round(cutoff2, 1)), x = cutoff2.i, y = cutoff2.a + 0.08) +
  annotate("text", label = paste("zero A1 infants\n", round(cutoff1, 1)), x = cutoff1.i + 0.08, y = cutoff1.a - 0.06) +
  labs(x = "Infant harvest proportion", y = "Adult harvest proportion", title = "ROC curve of adult and infant harvest proprtions") +
  theme_bw()

```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
flux::auc(x = (1 - prop.infants), y = (1 - prop.adults))

```


##### Section 10: (10 points) ###################

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population
 	
```{r Part_10} 	
matrix(
  c(cutoff0, round(cutoff0.a, 3), round(cutoff0.i, 3), round((cutoff0.a * total.adults + cutoff0.i * total.infants) / (total.adults + total.infants),3),
    cutoff1, round(cutoff1.a, 3), round(cutoff1.i, 3), round((cutoff1.a * total.adults + cutoff1.i * total.infants) / (total.adults + total.infants),3),
    cutoff2, round(cutoff2.a, 3), round(cutoff2.i, 3), round((cutoff2.a * total.adults + cutoff2.i * total.infants) / (total.adults + total.infants),3)), # the data elements 
  nrow = 3,              # number of rows 
  ncol = 4,              # number of columns 
  byrow = TRUE,        # fill matrix by rows 
  dimnames = list(
  c("max difference", "zero A1 infants", "equal error"),         # row names 
  c("Volume", "TPR", "FPR", "Prop. Yield"))) # column names 

## https://towardsdatascience.com/understanding-auc-roc-curve-68b2303cc9c5
```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the three cutoffs determined in this assignment.**   

***Answer: ROC is a plot of the true positive rate (TPR on the y-axis) against the false positive rate (FPR on the x-axis) for the different possible harvest cutoffs along with volume values. It tells how much model is capable of distinguishing between right amount harvests by adults vs infants. An excellent model has AUC near to the 1 which means it has a good measure of separability. A poor model has AUC near to the 0 which means it has the worst measure of separability. We have 0.87 AUC which has a good measure separability. There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. So, “Max difference” has the highest volume but lowest FPR (left side of the x-axis) and lowest proportion yield. “Zero A1 infants” has the lowest volume but highest FPR (right side of the x-axis) and highest proportion yield. Equal harvest is in between “Max difference” and “zero A1 infants.” ***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for implementation of a cutoff? 
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer: There is a tradeoff between risk and benefits in managing abalone harvest among three potential cutoffs or choices with the corresponding true positive rates, false positive rates, and total proportional yield; proportion harvested considering all abalone, adults, and infants. We would like to harvest infants with right cutoff choices with the right yield because we want to protect infant abalones since it represents future harvests. Not having information on sampling locations, sampling methods, mislabeling the ages or type, nutrition, predation, prior harvesting, disease, and other external factors can create limitations to our analysis. The confusion matrix we build focuses on true positive and false positive (type 1 error) rates and it is ignoring false negative and true negatives.  We should be cautious on overharvesting infant (to sustain future harvests) and adult (to sustain the population growth) abalones because there is false positive or type 1 error in our matrix and we need to minimize false positives.  We need to refine and train our model to reduce the error types when we predict thresholds on harvesting infant and adult abalone populations. We also need to validate if our sampling method is capturing overall abalone population correctly and labeling them right.***