---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Siddikov, Alisher"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("C:\\Users\\asidd\\Desktop\\MSDS\\401 Statistics with R\\5\\Assignment 1\\abalones.csv", sep = ",")


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections ##########################

##### Section 1: (6 points) Summarizing the data.

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

summary(mydata)

table(mydata$CLASS, mydata$RINGS)
#addmargins(table(mydata$CLASS, mydata$RINGS))
#View(round(describe(mydata),1))
```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: Variable types: SEX and CLASS are factors, RINGS is an integer, and the rest of the variables are numeric. By looking at the summary output and comparing mean values to the median along with their maximum and minimum values, we can see that most of the numeric variables are skewed except HEIGHT. We need to do further analysis to understand the outliers. We need to investigate outliers to understand if they are a legitimate part of the data or if there is some rational reason for excluding them from the analysis. Table: By looking at the frequency table there is a perfect correlation between age Class and Rings because age classification was based on Rings.***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
table1 <- table(mydata$SEX, mydata$CLASS)
addmargins(table1)

barplot(table1,beside = TRUE, col = c("darkred", "darkblue", "darkgreen"), 
        legend = c("Female", "Infant", "Male"),
        main = "CLASS membership, SEX - differentiated", 
        ylab = "Frequency", xlab = "Class")
```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: The frequency table and graph above tell us the distribution of the abalones by Sex (Male, Female, and Infants) and by age Class. The Class represents the age of the abalones by the number of rings in each with A1 being the youngest and A5 being the oldest. The number of infant abalones is almost the same as the number of female abalones. About 68% ((91+133)/329) of infant abalones represent younger age classes, A1 and A2. The number of infant abalones increase until about mid-age (A3) and declines after it. Interestingly 12% of infant abalones are classified A4 and A5 and it needs to be investigated The number of male and female abalones have the similar age distribution, however, the number of male abalones is slightly more than females. The number of female and male abalones start increasing from the youngest age class A1 to the mid-age class A3 and start decreasing as the age classification increases from mid-age class to the oldest age class. About 60% of male and female abalones represent mid-age classification, A3 and A4. Interestingly, about 73% (753/1036) all abalones represent around mid-age, from A2 to A4.***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
work <- mydata[sample(1:nrow(mydata), 200, replace=FALSE),]
plot(work[,2:6])

```

-----

##### Section 2: (5 points) Summarizing the data using graphics.

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
#ggplot(data = mydata, aes(x = VOLUME, y = WHOLE)) + 
#  geom_point(aes(color = CLASS),size = 1) + ggtitle("Whole weight, in grams | Volume, cm3")

plot(mydata$VOLUME,mydata$WHOLE, col = mydata$CLASS, pch = 16, cex = 0.75,lwd = 2,
     main = "Whole weight, in grams | Volume, cm3", xlab = "Volume", ylab = "Whole weight")
legend(x = 900, y = 100, legend=levels(mydata$CLASS), col = sort(unique(mydata$CLASS)), pch = 16, cex = 0.75,lwd = 2)
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
plot(mydata$WHOLE,mydata$SHUCK, col = mydata$CLASS, pch = 16, cex = 0.75,lwd = 2,
     main = "Shuck weight, in grams| Whole weight, in grams", xlab = "Whole Weight", ylab = "Shuck weight")
abline(a=0,b=max(mydata$SHUCK/mydata$WHOLE))
legend("bottomright", legend=levels(mydata$CLASS), col = sort(unique(mydata$CLASS)), pch = 16, cex = 0.75,lwd = 2)

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: Both charts are perfectly correlated. However, the first chart slightly dispersed than in the second chart.Younger class of abalones are closer to zero in both charts because they have fewer weights and smaller volumes. 1st scatter chart tries to show a relationship between the whole weight of abalone with shell and volume (length x diameter x height) of abalone with shell; it is color-coded by age class. As age class increases, the whole weight and volume get dispersed.2nd scatter chart tries to show a relationship between the whole weight of abalone with shell and shucked weight of abalone meat without a shell; it is color-coded by age class. According to the second chart, as class age increases, the shucked weight of abalones tend to decrease, it means that as abalones get older the shell weighs more than the meat. It is true because as an abalone gets older, shell growth is faster than shuck growth. In addition, the max ratio of shucked weight to whole weight is shown by a straight line, however, most of the abalones fall under this line which could indicate the shell weigh more than the meat and older abalones may not yield more meat.***

-----

##### Section 3: (8 points) Getting insights about the data using graphs.

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
par(mfrow = c(3,3))
hist(mydata$RATIO[mydata$SEX == "F"], col = "red", main = "Female RATIO", xlab ="")
hist(mydata$RATIO[mydata$SEX == "I"], col = "green", main = "Infant RATIO", xlab ="")
hist(mydata$RATIO[mydata$SEX == "M"], col = "blue", main = "Male RATIO", xlab ="")


boxplot(mydata$RATIO[mydata$SEX == "F"], col = "red", main = "Female RATIO")
boxplot(mydata$RATIO[mydata$SEX == "I"], col = "green", main = "Infant RATIO")
boxplot(mydata$RATIO[mydata$SEX == "M"], col = "blue", main = "Male RATIO")


qqnorm(mydata$RATIO[mydata$SEX == "F"], col = "red", main = "Female RATIO")
qqline(mydata$RATIO[mydata$SEX == "F"], col = "black")

qqnorm(mydata$RATIO[mydata$SEX == "I"], col = "green", main = "Infant RATIO")
qqline(mydata$RATIO[mydata$SEX == "I"], col = "black")

qqnorm(mydata$RATIO[mydata$SEX == "M"], col = "blue", main = "Male RATIO")
qqline(mydata$RATIO[mydata$SEX == "M"], col = "black")

par(mfrow = c(1,1))
```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: The ratio is derived from shuck weight divided by abalone volume. Histogram of ratios of all three sex classes was somewhat normally distributed, has a bell shape curve. We can also see that QQ plots were aligned to the straight line except for upper bound. Histograms and QQ plots suggest that the data skewed right, and boxplot suggests that the skewness is due to outliers. The female ratio has a more normal distribution than infant and male ratios. ***

(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}
f <- mydata$RATIO[mydata$SEX == "F"]
outliers_f <- boxplot.stats(f)$out
fo <- f[which(f %in% outliers_f)]
fo

i <- mydata$RATIO[mydata$SEX == "I"]
outliers_i <- boxplot.stats(i)$out
io <- i[which(i %in% outliers_i)]
io

m <- mydata$RATIO[mydata$SEX == "M"]
outliers_m <- boxplot.stats(m)$out
mo <- m[which(m %in% outliers_m)]
mo

ratio_outliers <- rbind(
mydata[mydata$SEX == "F",][f %in% outliers_f,],
mydata[mydata$SEX == "I",][i %in% outliers_i,],
mydata[mydata$SEX == "M",][m %in% outliers_m,])
kable(ratio_outliers)

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: We can see that there is a similar number of mild outliers (between 0.21 and 0.25) across male, female, and infant abalones. However, a female has one extreme outlier (0.31) and one very mild outlier (0.07).***

-----

##### Section 4: (8 points) Getting insights about possible predictors.

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}

par(mfrow = c(2,2))
boxplot(VOLUME ~ CLASS, data = mydata, xlab = "CLASS", ylab = "VOLUME")
boxplot(WHOLE ~ CLASS, data = mydata, xlab = "CLASS", ylab = "WHOLE")

plot(x = mydata$RINGS, y = mydata$VOLUME, xlab = "RINGS", ylab = "VOLUME")
plot(x = mydata$RINGS, y = mydata$WHOLE, xlab = "RINGS", ylab = "WHOLE")
```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: We can see that boxplots illustrate as age class increases from infant to early stage of adulthood, the volume and whole weight also increases. Infants gain whole weight and volume between age class A1 and A3 exponentially. Hence, the whole weight and volume are good at predicting infants (A1 and A2 age classes) versus adults (A3, A4, and A5 age classes). Adults from age class A3 to A5 are not good predictors of whole weight and volume because the median is not much different. Scatter plot tries to show a relationship between rings and volume/whole weight. We can also observe that the number of abalone population increase as rings increase, however, there is a denser abalone population until about rings 15. So, abalones expected life can be between 15 and 25 rings. Please note that more rings mean the older the age. ***

-----

##### Section 5: (12 points) Getting insights regarding different groups in the data.

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
v_mean <- aggregate(VOLUME ~ SEX + CLASS, data = mydata, mean)
s_mean <- aggregate(SHUCK ~ SEX + CLASS, data = mydata, mean)
r_mean <- aggregate(RATIO ~ SEX + CLASS, data = mydata, mean)


v_matrix <- matrix(data = round(v_mean[,3],2), nrow = 3, byrow = FALSE)
colnames(v_matrix) <- c("A1","A2","A3","A4","A5")
rownames(v_matrix) <- c("Female", "Infant", "Male")

s_matrix <- matrix(data = round(s_mean[,3],2), nrow = 3, byrow = FALSE)
colnames(s_matrix) <- c("A1","A2","A3","A4","A5")
rownames(s_matrix) <- c("Female", "Infant", "Male")

r_matrix <- matrix(data = round(r_mean[,3],4), nrow = 3, byrow = FALSE)
colnames(r_matrix) <- c("A1","A2","A3","A4","A5")
rownames(r_matrix) <- c("Female", "Infant", "Male")

kable(v_matrix, caption = "Volume")
kable(s_matrix, caption = "Shuck")
kable(r_matrix, caption = "Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}

interaction.plot(response = mydata$RATIO, trace.factor = mydata$SEX, x.factor = mydata$CLASS, fun = mean, main = "Mean Ratio per Class", 
                 trace.label = "Sex", xlab= "Class", ylab = "Ratio", type = "l", lty = 1, col= c("black", "blue", "red"))

interaction.plot(response = mydata$VOLUME, trace.factor = mydata$SEX, x.factor = mydata$CLASS, fun = mean, main = "Mean Volume per Class",
                 trace.label = "Sex", xlab= "Class", ylab = "Volume", type = "l", lty = 1, col= c("black", "blue", "red"))

interaction.plot(response = mydata$SHUCK, trace.factor = mydata$SEX, x.factor = mydata$CLASS, fun = mean, main = "Mean Weight per Class", 
                 trace.label = "Sex", xlab= "Class", ylab = "Shuck", type = "l", lty = 1, col= c("black", "blue", "red"))

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: The first chart tries to illustrate a relationship between the mean ratio and age class.  Since it is a ratio of shucked weight and volume, the first chart can summarize all three charts. Mean ratio increases when a shucked weight goes up, but the volume goes down. So, if an average ratio is higher, then there is more meat in a smaller sized (volume) abalone. We can see that there are more meat in adult (male and female) abalones than infants, however, as age increases there is less meat which means that less shucked weight but bigger volume.  Also, males gain more shuck weight between at younger age (from A1 to A2) but they lose shuck weight at a faster rate after A2. Females at a younger age start out much larger than the males but males catch up to the females around age class A2. On average, we can also see those female abalones have more meat and more volume than males. ***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
par(mfrow = c(2, 2))

a <- subset(mydata, mydata$RINGS < 16)

boxplot(a$VOLUME[a$SEX == "I"] ~ a$RINGS[a$SEX == "I"], ylim=c(0,1100), main= "Infant Volume | Rings", col= "blue", ylab = "Volume", Xlab = "Rings")
boxplot(a$VOLUME[a$SEX != "I"] ~ a$RINGS[a$SEX != "I"], ylim=c(0,1100), main= "Adult Volume | Rings", col= "red", ylab = "Volume", Xlab = "Rings")

boxplot(a$WHOLE[a$SEX == "I"] ~ a$RINGS[a$SEX == "I"], ylim=c(0,400), main= "Infant Whole Weight | Rings", col= "blue", ylab = "Whole Weight", Xlab = "Rings")
boxplot(a$WHOLE[a$SEX != "I"] ~ a$RINGS[a$SEX != "I"], ylim=c(0,400), main= "Adult Whole Weight | Rings", col= "red", ylab = "Whole Weight", Xlab = "Rings")

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: As adult abalones grow, they gain whole weight and volume faster until ring 5, they have steady gain until ring 10, and then their gain slows down. As infant abalones grow, they gain more in the beginning and then their gain fluctuates as they get older. We can also see that most of the volume and whole weight of abalones happen when they are adults. Adult abalones have longer whiskers and taller boxes, so their whole weight and volumes more dispersed than infants. Similarly, adults are more dispersed as the number of rings grows.***

-----

##### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).

**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: Controlling the outliers by introducing average and boxplots gave a deeper look at a relationship between age classes and ratios (volume and shuck weight). There might be errors in age classification because reading rings are difficult and time-consuming. I believe that infants should not be classified with adult ages and adult ring counts. Since there are difficulties in identifying rings, other measurements such as volume, shuck weight, and their ratios can be used to identify and classify their age. ***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: Answer: I would ask the following questions: <br> Distribution of the data and sampling errors: What sampling method was used and how was it obtained? How accurately is this sample represent the total population? How well the data distributed? How big are a population and sample data? Are they any skewness of the data? What do outliers mean and how should we interpret them? <br> Time factor and incidents: How long ago was the study conducted and is it still relevant? Does anything change after the study? After the study conducted, were there any major incidents (oil spills, disease spread) or natural disasters which can change the studied population? After the study conducted, are there any ecosystem changes such as warming ocean temperature or migration of other sea animals that either compete area or harm the studied creature? <br> Data modification and non-sampling errors: Are there any missing data, recording error, measurement error, input processing error, analysis error, response error, or other factors? Were there any data cleaning, alterations, additions, classifications/groupings, or modifications happened? ***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: In an observational study, we measure or survey members of a sample without trying to affect them. Like experiments, observational studies attempt to understand cause-and-effect relationships. We need to investigate the connection or correlation of two variables and find the true cause. Are both correlated variables related to a third, unknown factor? What is the third, unknown factor? Is the third factor hidden to sell the idea? We need to ask further follow-up questions: Who conducted or funded the study? Is the study biased by the researcher? What is the purpose of the study? Does the correlation indicate the reverse causation, indirect causation, coincidence or other types of causations? We should not jump to a conclusion when we find the connection or correlation between two variables.***
